---
title: "Assignment 2"
author: "Nishanth Gandhidoss"
date: "10/04/2017"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```


```{r installing necessary packages, include=FALSE}
# installing the packages
installNewPackage <- function(packageName) {
        if(packageName  %in% rownames(installed.packages()) == FALSE)
        {
                install.packages(packageName, repos = "http://cran.us.r-project.org", dependencies=TRUE)
        }
}


```


## Question 1 

### Section (a)


```{r Question1(a), message=FALSE}

```


## Section (b)


```{r Question1(b)}

```


### Section (c)


```{r Question1(c), include=FALSE}

```


## Question 2 

## Section (a)

```{r Question2(a)}

```


### Section (b)


```{r Question2(b)}

```

The missmap shows missing values in the dataset in white and observed values in red. The columm from lodging to leaf.halo are most likely to missing in the dataset. 

```{r Question2(b)ii}

```


### Section (c)


```{r Question2(c), include=FALSE}

```


## Question 3 

### Section (a)


```{r Question3(a)}

```


### Section (b)

```{r Question3(b), include=FALSE}

```


### Section (c)


```{r Question3(c)}
# Plot the correlation plot 
corrplot(cor(bbbDescr), order="hclust", tl.cex = 0.01)
```


*** End of Solution ***

### Appendix - Coding

\# Installing the packages

installNewPackage <- function(packageName) {

        if(packageName  %in% rownames(installed.packages()) == FALSE)

        {

                install.packages(packageName, repos = "http://cran.us.r-project.org", dependencies=TRUE)

        }

}

installNewPackage("ggplot2")

installNewPackage("lattice")

installNewPackage("mlbench")

installNewPackage("caret")

installNewPackage("grid")

installNewPackage("gridExtra")

installNewPackage("reshape2")

installNewPackage("e1071")

installNewPackage("plyr")

installNewPackage("corrplot")

installNewPackage("Amelia")

installNewPackage("reshape2")

installNewPackage("mice")

library(ggplot2)

library(lattice)

library(mlbench)

library(caret)

library(grid)

library(gridExtra)

library(reshape2)

library(e1071)

library(plyr)

library(corrplot)

library(Amelia)

library(reshape2)

library(mice)

#### Question 1

\# Getting the data

data(Glass)

\# Get the Predictors alone 

pred <- Glass[,-10]

\# Custom histogram function

gg_histogram <- function(col_name) {

    ggplot(data = pred, aes(pred[col_name])) + ggtitle(col_name) + xlab(col_name) +

        theme(plot.title = element_text(hjust = 0.5)) + geom_histogram(bins = 40)

}

\# Calling the user defined function

RI_hist <- gg_histogram("RI")

Na_hist <- gg_histogram("Na")

Mg_hist <- gg_histogram("Mg")

Al_hist <- gg_histogram("Al")

Si_hist <- gg_histogram("Si")

K_hist <- gg_histogram("K")

Ca_hist <- gg_histogram("Ca")

Ba_hist <- gg_histogram("Ba")

Fe_hist <- gg_histogram("Fe")

\# Mulitplot custome function

Reference http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {

  library(grid)

  \# Make a list from the ... arguments and plotlist

  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  \# If layout is NULL, then use 'cols' to determine layout
  
  if (is.null(layout)) {
  
    \# Make the panel
  
    \# ncol: Number of columns of plots
  
    \# nrow: Number of rows needed, calculated from \# of cols
  
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
  
                    ncol = cols, nrow = ceiling(numPlots/cols))
  
  }

 if (numPlots==1) {
 
    print(plots[[1]])

  } else {
 
    \# Set up the page
 
    grid.newpage()
 
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))


   \# Make each plot, in the correct location

    for (i in 1:numPlots) {

      \# Get the i,j matrix positions of the regions that contain this subplot

      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))

    }

  }

}

\# Calling the multiplot function

multiplot(RI_hist, Na_hist, Mg_hist, Al_hist, Si_hist, K_hist, Ca_hist, Ba_hist, Fe_hist, cols = 3)

\# Relationship between the Predictors

cormat <- round(cor(pred), 4)

melted_cormat <- melt(cormat, varnames = c("X", "Y"))

ggplot(data = melted_cormat, aes(x=X, y=Y, fill=value)) + geom_tile() + ggtitle("Relationship between the Predictors") +

        theme(plot.title = element_text(hjust = 0.5))
        
\# Outlier detection

par(mfrow = c(3, 3))

for(col_name in names(pred)[1:length(pred)]) {

    boxplot(pred[col_name], main = col_name, xlab = col_name)

}

\# Skewness

apply(pred, 2, skewness)

\# Box-Cox tranformation for the predictor skewness

boxcox_skewness = function(data) {

  box_cox_trans = BoxCoxTrans(data)

  data_BC = predict(box_cox_trans, data)

  skewness(data_BC) 

}

\# Remove the type or class label out of the dataset

Glass_predictors <- Glass[, -10]

apply(Glass_predictors, 2, boxcox_skewness)


#### Question 2

\# Getting the data

data(Soybean)

\# Set the screen layout for plotting

par(mfrow = c(3, 3))

\# Loop through the categorical variables of the data

for(col in names(Soybean)[2:10]) {

    barplot(table(Soybean[col]), main = paste("Bar chart for", col, "variable"), xlab = col, ylab = "Count", col = "grey")

}

for(col in names(Soybean)[11:19]) {

    barplot(table(Soybean[col]), main = paste("Bar chart for", col, "variable"), xlab = col, ylab = "Count", col = "grey")

}

for(col in names(Soybean)[20:28]) {

    barplot(table(Soybean[col]), main = paste("Bar chart for", col, "variable"), xlab = col, ylab = "Count", col = "grey")

}

for(col in names(Soybean)[29:dim(Soybean)[2]]) {

    barplot(table(Soybean[col]), main = paste("Bar chart for", col, "variable"), xlab = col, ylab = "Count", col = "grey")

}

\# Get categorical column numbers

\# For all categorical predictors, need to recall the data

SoybeanCat<-Soybean[, 2:dim(Soybean)[2]]

\# Calculating Near zero varaince

names(Soybean)[nearZeroVar(SoybeanCat)]

\# Missing map

missmap(Soybean, col = c("white", "red"))

\# Duplicate the data

Soybean1 <- Soybean

\# Check for NA across the row wise

Soybean1$has_nans_in_sample = apply(Soybean[,-1], 1, function(x){sum(is.na(x)) > 0 })

\# Tabulate the result

table(Soybean1[, c(1,37)])

\# Get the imputed data using mice

imputed <- mice(Soybean, m=1, maxit = 50, method = 'pmm', seed = 500)

imputed_data <- complete(imputed, 1)

\# Missing map

missmap(imputed_data, col = c("white", "red"))


#### Question 3

\# Loading the library

library(caret)

\# Data is loaded

data(BloodBrain)

bbbDescr[, c(1, 2)]

logBBB[1:5]

\# Look for degenerate columns

zero_cols = nearZeroVar(bbbDescr)

colnames(bbbDescr)[zero_cols]

\# Plot the correlation plot 

corrplot(cor(bbbDescr), order="hclust" )

\# Finding out which predictors to elliminate since they have too large correlations

highCorr = findCorrelation(cor(bbbDescr), cutoff=0.75 )

bbbDescr_independent = bbbDescr[,-highCorr]

\# Matrix has no values > cutoff=0.75

corrplot(cor(bbbDescr_independent))

\# Find the number of columns in the dataset before and after the doing correlation

ncol(bbbDescr)

ncol(bbbDescr_independent)

#### End of the Assignemnt